---
---
title: "APM1110: Formative Assessment 6"
subtitle: "R Programming and Probability Distributions"
author: "Ricafort, Nathalie Faith A."
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

# I. Geometric Distribution

The geometric distribution models the waiting time until the first success
in a sequence of independent Bernoulli trials. In this activity, the
probability of success on each trial is \(p = 0.2\).

R's `rgeom()` function generates the number of failures before the first
success. Since the problem asks for the number of trials required to achieve
the first success, 1 is added to every generated value.

## 1. Set the probability of success

```{r geometric-probability}
p <- 0.2
p
```

The probability of success in every trial is `r p`.

## 2. Generate 1,000 geometric random variables

A seed is set so that the same simulated sample and numerical results are
obtained whenever the document is rendered.

```{r geometric-simulation}
set.seed(1110)

n <- 1000

# rgeom() returns failures before the first success.
# Add 1 to obtain total trials required for the first success.
x <- rgeom(n = n, prob = p) + 1

# Display the first 20 simulated values.
head(x, 20)
```

A total of `r length(x)` observations were generated.

## 3. Calculate the basic statistics

```{r geometric-statistics}
mean_x <- mean(x)
var_x  <- var(x)
sd_x   <- sd(x)

mean_x
var_x
sd_x
```

For comparison, the theoretical statistics for the number of trials until
the first success are:

\[
E(X) = \frac{1}{p}
\]

\[
\operatorname{Var}(X) = \frac{1-p}{p^2}
\]

\[
\operatorname{SD}(X) = \sqrt{\frac{1-p}{p^2}}
\]

```{r geometric-theoretical}
theoretical_mean <- 1 / p
theoretical_variance <- (1 - p) / p^2
theoretical_sd <- sqrt(theoretical_variance)

theoretical_mean
theoretical_variance
theoretical_sd
```
## 4. Print the required formatted output

```{r geometric-output}
cat(
  "Number of trials required to achieve first success:",
  length(x),
  "\n"
)

cat(
  sprintf("Mean (in 2 decimal places): %.2f\n", mean_x)
)

cat(
  sprintf("Variance (in 2 decimal places): %.2f\n", var_x)
)

cat(
  sprintf("Standard deviation (in 2 decimal places): %.2f\n", sd_x)
)
```

The simulated statistics should be reasonably close to the theoretical mean
of `r sprintf("%.2f", theoretical_mean)`, variance of
`r sprintf("%.2f", theoretical_variance)`, and standard deviation of
`r sprintf("%.2f", theoretical_sd)`. Small differences are expected because
the values were obtained from a random simulation.

## 5. Plot the histogram

```{r geometric-histogram, fig.cap="Histogram of simulated trials required to achieve the first success."}
hist(
  x,
  breaks = seq(0.5, max(x) + 0.5, by = 1),
  main = "Geometric Distribution: Trials Until First Success",
  xlab = "Number of trials required",
  ylab = "Frequency",
  col = "steelblue",
  border = "white",
  xlim = c(0, max(x) + 1)
)
```

The histogram is right-skewed. Lower trial counts occur more frequently
because there is a chance of success on every trial. Large waiting times are
possible, but they become progressively less frequent.

# II. Hypergeometric Distribution

Suppose 10% of the IC chips in each box are defective. A sample of 10 chips
is selected without replacement.

Let \(X\) denote the number of defective chips in the sample. The phrase
"more than 10% defective" means:

\[
\frac{X}{10} > 0.10
\]

Therefore:

\[
X > 1
\]

Since \(X\) is an integer, the event is equivalent to \(X \ge 2\). Thus, the
required probability is:

\[
P(X > 1) = 1 - P(X \le 1)
\]

In R, the hypergeometric cumulative distribution function is:

```text
phyper(q, m, n, k)
```

where:

- `q` is the largest number of defectives included in the cumulative
  probability;
- `m` is the number of defective chips in the box;
- `n` is the number of non-defective chips in the box; and
- `k` is the sample size.

## 1. Sample of 10 from a box of 40

A box of 40 contains:

- \(0.10(40) = 4\) defective chips;
- \(40 - 4 = 36\) non-defective chips; and
- a sample size of 10 chips.

```{r hypergeometric-40}
box_size_40 <- 40
defective_40 <- 0.10 * box_size_40
nondefective_40 <- box_size_40 - defective_40
sample_size <- 10

# P(X > 1) = 1 - P(X <= 1)
prob_more_10_40 <- phyper(
  q = 1,
  m = defective_40,
  n = nondefective_40,
  k = sample_size,
  lower.tail = FALSE
)

cat(
  sprintf(
    "Probability for a sample of 10 from a box of 40: %.6f\n",
    prob_more_10_40
  )
)

cat(
  sprintf(
    "Percentage probability: %.2f%%\n",
    100 * prob_more_10_40
  )
)
```

As a manual check, the same result can be computed by adding the
probabilities for 2, 3, and 4 defective chips:

```{r hypergeometric-40-check}
x_values_40 <- 2:min(defective_40, sample_size)

probability_check_40 <- sum(
  dhyper(
    x_values_40,
    m = defective_40,
    n = nondefective_40,
    k = sample_size
  )
)

probability_check_40
```

**Interpretation:** The probability that more than 10% of the 10 sampled
chips are defective is approximately
`r sprintf("%.4f", prob_more_10_40)`, or
`r sprintf("%.2f", 100 * prob_more_10_40)`%.

## 2. Sample of 10 from a box of 5,000

A box of 5,000 contains:

- \(0.10(5000) = 500\) defective chips;
- \(5000 - 500 = 4500\) non-defective chips; and
- a sample size of 10 chips.

```{r hypergeometric-5000}
box_size_5000 <- 5000
defective_5000 <- 0.10 * box_size_5000
nondefective_5000 <- box_size_5000 - defective_5000

# P(X > 1) = 1 - P(X <= 1)
prob_more_10_5000 <- phyper(
  q = 1,
  m = defective_5000,
  n = nondefective_5000,
  k = sample_size,
  lower.tail = FALSE
)

cat(
  sprintf(
    "Probability for a sample of 10 from a box of 5000: %.6f\n",
    prob_more_10_5000
  )
)

cat(
  sprintf(
    "Percentage probability: %.2f%%\n",
    100 * prob_more_10_5000
  )
)
```

**Interpretation:** The probability that more than 10% of the 10 sampled
chips are defective is approximately
`r sprintf("%.4f", prob_more_10_5000)`, or
`r sprintf("%.2f", 100 * prob_more_10_5000)`%.

## Comparison of the two cases

```{r hypergeometric-comparison}
comparison <- data.frame(
  Box_Size = c(40, 5000),
  Defective_Chips = c(defective_40, defective_5000),
  Nondefective_Chips = c(nondefective_40, nondefective_5000),
  Sample_Size = c(sample_size, sample_size),
  Probability = c(prob_more_10_40, prob_more_10_5000),
  Percentage = c(
    100 * prob_more_10_40,
    100 * prob_more_10_5000
  )
)

knitr::kable(
  comparison,
  digits = c(0, 0, 0, 0, 6, 2),
  col.names = c(
    "Box size",
    "Defective",
    "Non-defective",
    "Sample size",
    "Probability",
    "Percentage"
  ),
  caption = "Probability that more than 10% of the sampled chips are defective"
)
```

The probability for the larger box is close to the corresponding binomial
probability because sampling 10 chips has very little effect on a population
of 5,000. In the box of 40, sampling without replacement produces a stronger
finite-population effect.

# Conclusion

The geometric simulation models the total number of trials needed to obtain
the first success when the success probability is 0.2. The simulated
statistics are close to the theoretical geometric-distribution values.

For the hypergeometric problems, "more than 10% defective" in a sample of 10
means at least 2 defective chips. The required probabilities were calculated
using the upper tail of the hypergeometric distribution.
